Recursive Solutions to First-Order Model Counting

### 2022-03-12 notes/notes.tex
First few slides for the new talk

\paragraph{Things I might need to explain.}
\begin{itemize}
\item atom, (positive/negative) literal, constant, predicate, variable, clause, unit clause, first-order logic
\item $\WMC$, $\wwp$, $\wwn$.
\item two parts: compilation and inference.
\item During inference, there is a domain size map $\size\colon \mathcal{D} \to \mathbb{N}_0$.
\item mention which rules are in $\Gamma$ and which ones are in $\Delta$ (and why tryCache has to be in $\Delta$). TODO: having notation for $\Delta$ and $\Gamma$ isn't that necessary
\item \textsc{ForcLift}
\item $\sqcup$ for both sets and functions
\end{itemize}

@@

TODO: Note: we will reuse this several times.
\begin{example} \label{example:first}
  Let $\phi \coloneqq \{\, c_1, c_2 \,\}$ be a formula, where
  \[
  c_1 \coloneqq (\{\, \neg p(X, Y), \neg p(X, Z) \,\}, \{\, (Y, Z) \,\}, \{\, X \mapsto a, Y \mapsto b, Z \mapsto b \,\}),
  \]
  and
  \[
  c_2 \coloneqq (\{\, \neg p(X, Y), \neg p(Z, Y) \,\}, \{\, (X, Z) \,\}, \{\, X \mapsto a, Y \mapsto b, Z \mapsto a \,\}),
  \]
  for some predicate $p$, variables $X$, $Y$, $Z$, and domains $a$ and $b$. Then in first-order logic $\phi$ could be read as
  \begin{align*}
    (\forall X \in a. \forall Y \in b. \forall Z \in b. Y \ne Z &\implies \neg p(X, Y) \lor \neg p(X, Z)) \land \\
    (\forall X \in a. \forall Y \in b. \forall Z \in a. X \ne Z &\implies \neg p(X, Y) \lor \neg p(Z, Y)).
  \end{align*}
\end{example}

@@

\paragraph{TODO}
\begin{itemize}
\item update the example to not refer to things that don't exist anymore
\item introduce/describe and \cref{alg:trycache} and describe the cache that's being used.
\item explain why $\rho \cup \gamma$ is possible
\item explain what the second return statement is about and why a third one is not necessary
\item explain the yield keyword
\item in the example below: write down both formula using the ForcLift format
\end{itemize}

@@

Diagramatically, \texttt{constructDomainMap} attempts to find $\gamma\colon \Doms(c) \to \Doms(d)$ s.t.
\[
\begin{tikzcd}
  \Vars(c) \ar[r, tail, two heads, "\beta"] \arrow[d, swap, "\delta_c"] & \Vars(d) \ar[d, "\delta_d"] \\
  \Doms(c) \ar[r, dashed, "\gamma"] \ar[d, hookrightarrow] & \Doms(d) \ar[d, hookrightarrow] \\
  \mathcal{D} \ar[r, swap, "\rho"] & \mathcal{D}.
\end{tikzcd}
\]
commutes (and returns \texttt{null} if such a function does not exist).

TODO: update this example. Fix references.
\begin{example} \label{example}
  Let $\phi$ be the formula from \cref{example:first} and $\psi \coloneqq \phi[\id, \{\, a \mapsto a', b \mapsto b^\bot \,\}]$ (i.e., $\phi$ with both of its domains replaced).

  Since $\#\phi = \#\psi$, and the formulas are non-empty, the algorithm proceeds with the for-loops on \cref{line:for1,line:for2,line:generateMaps}. Suppose $c$ in the algorithm refers to \cref{phi1}, and $d$ to \cref{psi1}. Since both clauses have three variables, in the worst case, function \texttt{generateMaps} would have $3!=6$ bijections to check. Suppose the identity bijection is picked first. Then \texttt{constructDomainMap} is called with the following parameters:
  \begin{itemize}
  \item $V = \{\, X, Y, Z \,\}$,
  \item $\delta_c = \{\, X \mapsto a, Y \mapsto b, Z \mapsto b \,\}$,
  \item $\delta_d = \{\, X \mapsto a', Y \mapsto b^\bot, Z \mapsto b^\bot \,\}$,
  \item $\beta = \{\, X \mapsto X, Y \mapsto Y, Z \mapsto Z \,\}$,
  \item $\rho = \emptyset$.
  \end{itemize}
  Since $\delta_i(Y) = \delta_i(Z)$ for $i \in \{\, c, d \,\}$, \texttt{constructDomainMap} returns $\gamma = \{\, a \mapsto a', b \mapsto b^\bot \,\}$. Thus, \texttt{generateMaps} yields its first pair of maps $(\beta, \gamma)$ to \cref{line:generateMaps}. Furthermore, the pair satisfies $c[\beta, \gamma] = d$.

@@

  Again we have two non-empty formulas with equal hash codes, so \texttt{generateMaps} is called with $c$ set to \cref{phi2}, $d$ set to \cref{psi2}, and $\rho = \{\, a \mapsto a', b \mapsto b^\bot \,\}$. Suppose \cref{line:bijection} picks the identity bijection first again. Then \texttt{constructDomainMap} is called with the following parameters:
  \begin{itemize}
  \item $V = \{\, X, Y, Z \,\}$,
  \item $\delta_c = \{\, X \mapsto a, Y \mapsto b, Z \mapsto a \,\}$,
  \item $\delta_d = \{\, X \mapsto a', Y \mapsto b^\bot, Z \mapsto a' \,\}$,
  \item $\beta = \{\, X \mapsto X, Y \mapsto Y, Z \mapsto Z \,\}$,
  \item $\rho = \{\, a \mapsto a', b \mapsto b^\bot \,\}$.
  \end{itemize}
  Since $\beta$ and $\rho$ `commute' (TODO: as in the diagram above), and there are no new domains in $\Doms(c)$ and $\Doms(d)$, $\gamma$ exists and is equal to $\rho$. Again, the returned pair $(\beta, \gamma)$ satisfies $c[\beta, \gamma] = d$. This $\gamma$ passes the \texttt{traceAncestors} checks exactly the same way as the one before, and \cref{line:recursion} calls \texttt{identifyRecursion($\emptyset$, $\emptyset$, $\rho$, true)}, which immediately returns $\rho = \{\, a \mapsto a', b \mapsto b^\bot \,\}$ as the final answer. This means that one can indeed use an FCG for $\WMC(\phi)$ to compute $\WMC(\psi)$ by replacing every mention of $a$ with $a'$ and every mention of $b$ with $b^\bot$.
\end{example}

>>>>

\paragraph{Things I might need to explain.}
\begin{itemize}
\item atom, (positive/negative) literal, constant, predicate, variable, clause, unit clause, first-order logic
\item $\WMC$, $\wwp$, $\wwn$.
\item two parts: compilation and inference.
\item During inference, there is a domain size map $\size\colon \mathcal{D} \to \mathbb{N}_0$.
\item mention which rules are in $\Gamma$ and which ones are in $\Delta$ (and why tryCache has to be in $\Delta$). TODO: having notation for $\Delta$ and $\Gamma$ isn't that necessary
\item My formalisation of constraints is more restrictive than the original (e.g., in the thesis).
\item \textsc{ForcLift}
\item $\sqcup$ for both sets and functions
\end{itemize}

@@

TODO: Note: we will reuse this several times.
\begin{example} \label{example:first}
  Let $\phi \coloneqq \{\, c_1, c_2 \,\}$ be a formula, where
  \begin{equation} \label{phi1}
    c_1 \coloneqq (\{\, \neg p(X, Y), \neg p(X, Z) \,\}, \{\, (Y, Z) \,\}, \{\, X \mapsto a, Y \mapsto b, Z \mapsto b \,\}),
  \end{equation}
  and
  \begin{equation} \label{phi2}
    c_2 \coloneqq (\{\, \neg p(X, Y), \neg p(Z, Y) \,\}, \{\, (X, Z) \,\}, \{\, X \mapsto a, Y \mapsto b, Z \mapsto a \,\}),
  \end{equation}
  for some predicate $p$, variables $X$, $Y$, $Z$, and domains $a$ and $b$. Then in first-order logic $\phi$ could be read as
  \begin{align*}
    (\forall X \in a. \forall Y \in b. \forall Z \in b. Y \ne Z &\implies \neg p(X, Y) \lor \neg p(X, Z)) \land \\
    (\forall X \in a. \forall Y \in b. \forall Z \in a. X \ne Z &\implies \neg p(X, Y) \lor \neg p(Z, Y)).
  \end{align*}
\end{example}

@@

\paragraph{TODO}
\begin{itemize}
\item introduce/describe \cref{alg:trycache} and describe the cache that's being used.
\item explain why $\rho \cup \gamma$ is possible
\item explain what the second return statement is about and why a third one is not necessary
\item explain the yield keyword
\item in the example below: write down both formula using the ForcLift format
\end{itemize}

@@

Diagramatically, \texttt{constructDomainMap} attempts to find $\gamma\colon \Doms(c) \to \Doms(d)$ s.t.
\begin{equation} \label{eq:commute}
\begin{tikzcd}
  \Vars(c) \ar[r, tail, two heads, "\beta"] \arrow[d, swap, "\delta_c"] & \Vars(d) \ar[d, "\delta_d"] \\
  \Doms(c) \ar[r, dashed, "\gamma"] \ar[d, hookrightarrow] & \Doms(d) \ar[d, hookrightarrow] \\
  \mathcal{D} \ar[r, swap, "\rho"] & \mathcal{D}.
\end{tikzcd}
\end{equation}
commutes (and returns \texttt{null} if such a function does not exist).

TODO: update this example. Fix references.
\begin{example} \label{example}
  Let $\phi$ be the formula from \cref{example:first} and $\psi \coloneqq \phi[\id, \{\, a \mapsto a', b \mapsto b^\bot \,\}]$ (i.e., $\phi$ with both of its domains replaced). Since $\#\phi = \#\psi$, and both formulas are non-empty, the algorithm proceeds with the for-loops on \cref{line:for1,line:for2,line:generateMaps}. Suppose $c$ in the algorithm refers to \cref{phi1}, and $d$ to \cref{psi1}. Since both clauses have three variables, in the worst case, function \texttt{generateMaps} would have $3!=6$ bijections to check. Suppose the identity bijection is picked first. Then \texttt{constructDomainMap} is called with the following parameters:
  \begin{itemize}
  \item $V = \{\, X, Y, Z \,\}$,
  \item $\delta_c = \{\, X \mapsto a, Y \mapsto b, Z \mapsto b \,\}$,
  \item $\delta_d = \{\, X \mapsto a', Y \mapsto b^\bot, Z \mapsto b^\bot \,\}$,
  \item $\beta = \{\, X \mapsto X, Y \mapsto Y, Z \mapsto Z \,\}$,
  \item $\rho = \emptyset$.
  \end{itemize}
  Since $\delta_i(Y) = \delta_i(Z)$ for $i \in \{\, c, d \,\}$, \texttt{constructDomainMap} returns $\gamma = \{\, a \mapsto a', b \mapsto b^\bot \,\}$. Thus, \texttt{generateMaps} yields its first pair of maps $(\beta, \gamma)$ to \cref{line:generateMaps}. Furthermore, the pair satisfies $c[\beta, \gamma] = d$.

@@

  Again we have two non-empty formulas with equal hash codes, so \texttt{generateMaps} is called with $c$ set to \cref{phi2}, $d$ set to \cref{psi2}, and $\rho = \{\, a \mapsto a', b \mapsto b^\bot \,\}$. Suppose \cref{line:bijection} picks the identity bijection first again. Then \texttt{constructDomainMap} is called with the following parameters:
  \begin{itemize}
  \item $V = \{\, X, Y, Z \,\}$,
  \item $\delta_c = \{\, X \mapsto a, Y \mapsto b, Z \mapsto a \,\}$,
  \item $\delta_d = \{\, X \mapsto a', Y \mapsto b^\bot, Z \mapsto a' \,\}$,
  \item $\beta = \{\, X \mapsto X, Y \mapsto Y, Z \mapsto Z \,\}$,
  \item $\rho = \{\, a \mapsto a', b \mapsto b^\bot \,\}$.
  \end{itemize}
  Since $\beta$ and $\rho$ commute (as in \cref{eq:commute}), and there are no new domains in $\Doms(c)$ and $\Doms(d)$, $\gamma$ exists and is equal to $\rho$. Again, the returned pair $(\beta, \gamma)$ satisfies $c[\beta, \gamma] = d$. \Cref{line:recursion} calls \texttt{identifyRecursion($\emptyset$, $\emptyset$, $\rho$, true)}, which immediately returns $\rho = \{\, a \mapsto a', b \mapsto b^\bot \,\}$ as the final answer. This means that one can indeed use an FCG for $\WMC(\phi)$ to compute $\WMC(\psi)$ by replacing every mention of $a$ with $a'$ and every mention of $b$ with $b^\bot$.
\end{example}
